---
title: "index"
author: "Alexander Rogers"
date: "2025-02-27"
output: html_document
---

```{r}
library(tidyverse)
library(languageR)
library(untidydata)
```

```{r}
data(package = "untidydata")
```

```{r}
spir <- spirantization
```

```{r}
#| label: Tidying the data.
sprn <- spir |>
  separate(
    col = id,
    into = c("prof", "num"),
    sep = "_",
    remove = FALSE
  ) |>
  pivot_longer(
    cols = c("cIntensity", "vIntensity"),
    names_to = "item",
    values_to = "intensity"
  ) |>
  pivot_wider(
    names_from = "position",
    values_from = "intensity"
  ) |>
  mutate(
    item = if_else(item == "cIntensity", "Obstruent", "Vowel")
  )
```

```{r}
#| label: descriptive statistics, Average intensity for each proficiency group.
sprn |>
  group_by(prof, item) |>
  summarize(
    avgin = mean(initial),
    avgvcv = mean(vcv),
    sdin = sd(initial),
    sdvcv = sd(vcv)
  )
```

```{r}
#| label: Fitting linear model

mod <- lm(
  formula = initial ~ vcv,
  data = sprn
)

print(mod)
```

```{r}
#| label: Vizualising linear relationships in data

sprn |>
  ggplot() +
  aes(x = initial, y = vcv, color = prof) +
  geom_point() +
  geom_smooth(method = "lm")
```

There appears to be a significant difference between proficiency groups. Native speakers have overal higher intensity than advanced speakers, who have overall higher intensity than beginners. Furthermore, it appears that if a speaker has higher intensities after an obstruent, they also have higher intensity following vowels, though this difference is blurrier between the learner groups, whereas native speakers have a clear pattern. 
